\section{Object Mutation and Aliasing}
\label{sec:object-mutation-and-aliasing}
A limitation that we have purposefully imposed on our object/class design is the inability to modify the values of instance variables. Value mutation is a foreign concept in some programming languages, but we have made extensive use of it throughout our time in the land of Java. In this section, we will discuss the implications of allowing instance variable mutation and how it can lead to some unexpected behavior.

To access a \ttt{private} instance variable, we design a \ttt{public} accessor method, which returns the instance variable. To modify a \ttt{private} instance variable, we design a \ttt{public} mutator method\index{mutator method}\index{setter method}, which takes in a parameter and assigns the instance variable to the parameter. Let's return to the \ttt{Point} class to demonstrate. Suppose that we instantiate a point $p$ to $(7, 4)$, but we then want to change or modify either coordinate. We can do so by calling the \ttt{setX} or \ttt{setY} methods, respectively. Testing setter methods is important to verify that a change occurred when invoking the setter/mutator method, which we confirm through the accessor method. Another way of phrasing this is that, when testing a mutator, we care about the \emph{side-effect}\index{side-effect} of the method rather than what it returns, namely nothing. Setter methods, or methods that modify outside values or data are definitionally \emph{impure}\index{impure}. Because we want to be able to alter an instance variable, these can no longer be marked as \ttt{final}, so we remove this keyword.\footnote{This is not to suggest that we should never use \ttt{final} instance variables. In fact, we \emph{should} use \ttt{final} instance variables whenever possible, since object mutation introduces the possibility of easy-to-overlook bugs.}

\begin{cl}[]{Testing \ttt{Point} Setter Methods}
\begin{lstlisting}[language=MyJava]
import static Assertions.assertAll;
import static Assertions.assertEquals;

class PointTester {

  @Test
  void testSetX() {
    Point p = new Point(7, 4);
    p.setX(3);
    assertEquals(3, p.getX());
  }

  @Test
  void testSetY() {
    Point p = new Point(7, 4);
    p.setY(2);
    assertEquals(2, p.getY());
  }
}
\end{lstlisting}
\end{cl}

\begin{cl}[]{\ttt{Point} Class with Accessor and Mutator Methods}
\begin{lstlisting}[language=MyJava]
class Point {

  private int x;
  private int y;

  public Point(int x, int y) { 
    this.x = x; 
    this.y = y; 
  }

  public int getX() { return this.x; }

  public int getY() { return this.y; }

  public void setX(int x) { this.x = x; }

  public void setY(int y) { this.y = y; }
}
\end{lstlisting}
\end{cl}

What are some consequences to mutating an object? One comes through the notion of \emph{object aliasing}. Recall that objects point to references in memory. Therefore if we instantiate a \ttt{Point} $p_1$, then initialize another \ttt{Point} $p_2$ to reference $p_1$, then both objects refer to the same \ttt{Point} instance in memory. If we modify $p_1$ through a setter method, then $p_2$ will reflect the change.

\begin{cl}[]{Object Aliasing}
\begin{lstlisting}[language=MyJava]
import static Assertions.assertAll;
import static Assertions.assertEquals;

class PointTester {

  @Test
  void pointAliasing() {
    Point p1 = new Point(7, 4);
    Point p2 = p1;
    p1.setX(3);
    assertEquals(3, p1.getX());
    assertEquals(3, p2.getX());
    assertEquals(p1, p2);
  }

  @Test
  void testSetY() {
    Point p1 = new Point(7, 4);
    Point p2 = p1;
    p1.setY(2);
    assertEquals(2, p1.getY());
    assertEquals(2, p2.getY());
    assertEquals(p1, p2);
  }
}
\end{lstlisting}
\end{cl}

This idea carries over to other, more complex classes as well. For example, strings, arrays, lists, and others are all objects, and therefore, they are all subject to object aliasing. Modifying one \ttt{ArrayList} instance will modify all other \ttt{ArrayList} instances that reference the same object. This is a common source of bugs in Java programs, and it is important to be aware of this behavior. In the following example, we will demonstrate aliasing through the \ttt{ArrayList} data structure containing \ttt{Point} objects. We add a series of \ttt{Point} instances to an \ttt{ArrayList}, which is then aliased by another \ttt{ArrayList}. We then add another \ttt{Point} instance to the first \ttt{ArrayList}, followed by a verification that the lists are the same size. Additionally, we traverse over the lists and verify that the elements are the same through the \ttt{==} operator. Remember that \ttt{==} returns whether or not two objects reference the same instance in memory. Because these lists are merely aliases of each other, they will contain references to the same \ttt{Point} instances.

\begin{cl}[]{Testing \ttt{ArrayList} Aliasing}
\begin{lstlisting}[language=MyJava]
import static Assertions.assertAll;
import static Assertions.assertEquals;

class PointTester {

  private final Point P1 = new Point(7, 4);
  private final Point P2 = new Point(3, 2);
  private final Point P3 = new Point(1, 8);

  @Test
  void testPointArrayListAliasing() {
    List<Point> list1 = new ArrayList<>(List.of(P1, P2, P3));
    List<Point> list2 = list1;
    list1.add(new Point(5, 6));

    // First we can verify that the lists are actually the same.
    assertTrue(list1 == list2);

    // Size testing.
    assertTrue(list1.size() == list2.size());

    // Make sure both lists contain the same elements.
    for (int i = 0; i < list1.size(); i++) {
      assertTrue(list1.get(i) == list2.get(i));
    }
  }
}
\end{lstlisting}
\end{cl}

\example{Now that we have classes, mutation, and accessibility, we can finally implement generic data structures such as an \ttt{ArrayList}. In this example we will implement the behavior of the \ttt{ArrayList} data structure, which means we finally understand what is going on under the hood in the Collections API. Let's design a class \ttt{MiniArrayList}, which operates over any type using generics. Like generic static methods, we must quantify the generic type, but unlike static methods, however, we quantify the type over the class declaration, meaning that all instance and static methods observe/respect the quantifier and do not need to be separately quantified.}

In addition to the class header, what else does an \ttt{ArrayList} store? A backing array of elements and its corresponding length, of course! The array, as we described in Chapter~\ref{chapter-arrays-collections}, dynamically resizes as we add or insert elements. The logical size of the array, i.e., the number of presently-existing elements, is stored in \ttt{size}, whereas the current capacity, i.e., how many elements can currently be stored without a resize, is stored in \ttt{capacity}. Our class will provide two constructors: one that instantiates the backing array to store ten elements, and another that allows the user to specify. Interestingly, this shows off a great example of one constructor calling another of the same class, an idea called \emph{constructor chaining}. 

\begin{cl}[]{\ttt{MiniArrayList} Class}
\begin{lstlisting}[language=MyJava]
class MiniArrayList<T> {

  private T[] elements;
  private int size;
  private int capacity;

  public MiniArrayList() { this(10); }

  public MiniArrayList(int capacity) {
    this.size = 0;
    this.capacity = capacity;
  }
}
\end{lstlisting}
\end{cl}

Notice that we declare an array of type \ttt{T} to store the elements of our mini array list. We now must instantiate the array inside the second constructor. The problem is that we cannot instantiate an array of a generic type. Recall that generics are a compile-time feature, and arrays are a runtime feature. Therefore, we cannot instantiate an array of a generic type. Instead, we must instantiate an array containing type \ttt{Object}, followed by a cast to contain type \ttt{T}.\footnote{To be pedantic, the array is of type \ttt{Object[]}, and we cast it to type \ttt{T[]}.} This is called an \emph{unchecked cast}, and it is a necessary evil in Java to support powerful classes that operate over generic arrays.

\begin{cl}[]{\ttt{MiniArrayList} Class Instantiating the Backing Array}
\begin{lstlisting}[language=MyJava]
class MiniArrayList<T> {

  private T[] elements;
  private int size;
  private int capacity;

  public MiniArrayList() { this(10); }

  public MiniArrayList(int capacity) {
    this.size = 0;
    this.capacity = capacity;
    this.elements = (T[]) new Object[capacity];
  }
}
\end{lstlisting}
\end{cl}

We now need to implement the \ttt{add} method, which adds an element to the end of the array list. We first check if the array is full, and if so, we resize the array. We then add the element to the end of the array and increment the size. Resizing the array is, fortunately, not complicated; all we need to do is instantiate a new, larger array, copy the existing elements over, then reassign the instance variable. The question now is, by what factor should the array capacity increase? This decision is implementation-dependent, but we will use a doubling factor out of convenience. We make \ttt{resize} private because it is an implementation detail that the user does not need to know about. To write coherent tests, we should also write the \ttt{get} method, which returns the element at a given index, as well as \ttt{size}, which returns the number of logical elements in the list. For now, we will not worry about bounds checking, but we will return to this in a later chapter on exceptions.

\begin{cl}[]{Testing \ttt{MiniArrayList} \ttt{add} Method}
\begin{lstlisting}[language=MyJava]
import static Assertions.assertAll;
import static Assertions.assertEquals;

class MiniArrayListTester {

  @Test
  void testAdd() {
    MiniArrayList<Integer> list = new MiniArrayList<>();
    list.add(100);
    list.add(200);
    list.add(300);
    assertAll(
      () -> assertEquals(3, list.size()),
      () -> assertEquals(100, list.get(0)),
      () -> assertEquals(200, list.get(1)),
      () -> assertEquals(300, list.get(2)));
  }
}
\end{lstlisting}
\end{cl}

\begin{cl}[]{\ttt{MiniArrayList} Adding Elements}
\begin{lstlisting}[language=MyJava]
class MiniArrayList<T> {

  private static final int RESIZE_FACTOR = 2;

  /**
   * Adds an element to the end of the list.
   * @param element the element to add.
   */
  public void add(T element) {
    if (this.size == this.capacity) { this.resize(); }
    this.elements[this.size++] = element;
  }

  /**
   * Resizes the backing array by a factor specified by the class.
   */
  private void resize() {
    this.capacity *= RESIZE_FACTOR;
    T[] newArray = (T[]) new Object[this.capacity];
    for (int i = 0; i < this.size; i++) { newArray[i] = this.elements[i]; }
    this.elements = newArray;
  }

  public T get(int index) { return this.elements[index]; }

  public int size() { return this.size; }
}
\end{lstlisting}
\end{cl}

We will write two more methods: \ttt{insert} and \ttt{remove}, which inserts an element $e$ at a given index $i$, and removes an element $e$ respectively. These two methods are similar in that they alter the backing array by shifting its values right and left. Accordingly, our implementation will contain the private helper methods \ttt{shiftRight} and \ttt{shiftLeft}. If we attempt to insert an element into a list that must be resized, we call \ttt{resize}. Both \ttt{insert} and \ttt{remove} warrant test cases! Like the \ttt{get} counterpart, neither of these new methods will perform bounds checking, so testing out-of-bounds behavior, for the time being, is not pertinent.

\begin{cl}[]{Testing \ttt{MiniArrayList} \ttt{insert} and \ttt{remove} Methods}
\begin{lstlisting}[language=MyJava]
import static Assertions.assertAll;

class MiniArrayListTester {

  @Test
  void testInsert() {
    MiniArrayList<Integer> list = new MiniArrayList<>();
    list.add(100);
    list.add(300);
    list.insert(1, 150);
    assertAll(
      () -> assertEquals(3, list.size()),
      () -> assertEquals(100, list.get(0)),
      () -> assertEquals(150, list.get(1)),
      () -> assertEquals(300, list.get(2)));
  }

  @Test
  void testRemove() {
    MiniArrayList<Integer> list = new MiniArrayList<>();
    list.add(100);
    list.add(300);
    list.remove(0);
    assertAll(
      () -> assertEquals(1, list.size()),
      () -> assertEquals(300, list.get(0)));
  }
}
\end{lstlisting}
\end{cl}

\begin{cl}[]{\ttt{MiniArrayList} Inserting and Removing Elements}
\begin{lstlisting}[language=MyJava]
class MiniArrayList<T> {

  // ... other details not shown.

  /**
   * Inserts an element at the given index.
   * @param e the element to insert.
   * @param idx the index to insert at.
   */
  public void insert(T e, int idx) {
    if (this.size == capacity) { this.resize(); }
    this.shiftRight(idx);
    this.elements[idx] = e;
  }

  /**
   * Removes the element at the given index.
   * @param idx the index to remove.
   * @return the element removed.
   */
  public T remove(int idx) {
    T e = this.get(idx);
    this.shiftLeft(idx);
    this.size--;
    return e;
  }
}
\end{lstlisting}
\end{cl}

\begin{cl}[]{Shifting Left and Right}
\begin{lstlisting}[language=MyJava]
class MiniArrayList<T> {

  // ... other details not shown.

  /**
   * Shifts all elements to the left of the given index one position leftwards.
   * Note that this method overwrites the element at the given index.
   * @param idx the index to shift left of.
   */
  private void shiftLeft(int idx) {
    for (int i = idx; i < this.size - 1; i++) {
      this.elements[i] = this.elements[i + 1];
    }
  }

  /**
   * Shifts all elements to the right of the given index one position rightwards.
   * @param idx the index to shift right of.
   */
  private void shiftRight(int idx) {
    for (int i = size - 1; i > idx; i--) {
      this.elements[i] = this.elements[i - 1];
    }
  }
}
\end{lstlisting}
\end{cl}

\example{Let's see a few more examples of object aliasing and mutation. These examples will not be meaningful in what they represent, but are great exercises in testing your understanding. We will create five classes: \ttt{A}, \ttt{B}, \ttt{C}, \ttt{D}, and \ttt{E}. Class \ttt{A} contains one mutable string instance variable; its constructor assigns the instance variable to the parameter thereof. Classes \ttt{B} and \ttt{C} are identical aside from the name: they contain an immutable object of type \ttt{A} as an instance variable. Class \ttt{D} stores an integer array of ten elements. Finally, class \ttt{E} stores a mutable integer as an instance variable.}

We present several test cases that assert different pieces of these classes. We will analyze each one and determine why it uses either \ttt{assertEquals} or \ttt{assertNotEquals} in its comparison. Our first series of tests only focuses on classes \ttt{A}, \ttt{B}, and \ttt{C} to keep things simple. We insert blanks in the assertion statements for you to fill in as exercise before checking your answers.

\begin{cl}[]{Testing Object Mutation and Aliasing}
\begin{lstlisting}[language=MyJava]
import static Assertions.assertAll;
import static Assertions.assertEquals;

class ClassTester {

  @Test
  void testOne() {
    final A a = new A("Hello!");
    B b = new B(a);
    C c = new C(a);

    // What do each of these output?
    assert______(b.getA(), c.getA());
    assert______(b.getA().getS(), c.getA().getS());
    a.setS("Hi!");
    assert______(b.getA().getS(), c.getA().getS());
    b.getA().setS("howdy!");
    assert______(b.getA().getS(), c.getA().getS());

    B b2 = new B(a);
    assert______(a, b2.getA());
    assert______(b, b2);
    b = b2;
    assert______(a, b.getA());
    assert______(b, b2);
  }
}
\end{lstlisting}
\end{cl}

To set the scene, we first declare \ttt{a} as an immutable instance of \ttt{A} with the string literal \ttt{"Hello!"}. Then, we instantiate objects \ttt{b} and \ttt{c} of types \ttt{B} and \ttt{C} respectively, each receiving \ttt{a} as an argument to their constructors. 

Comparing \ttt{b.getA()} against \ttt{c.getA()} is a comparison of two references to the same object. Because \ttt{a} is immutable, we cannot change its value, so both \ttt{b} and \ttt{c} will always refer to the same object. Therefore, we use \ttt{assertEquals} to compare the two references. In particular, passing \ttt{a} to both constructors passes a reference to the same object.\footnote{In Figure~\ref{fig:memoryref}, we use memory addresses to refer to the location of \ttt{a}. To be a bit pedantic, objects are not stored directly in system memory per se, but rather a location accessible by the Java Virtual Machine.}
\begin{figure}[H]
\centering
\begin{tikzpicture}
  % Object a
  \node[rectangle, draw] (a) {\texttt{a = 0xffff2345}};

  % References b and c
  \node[above left=of a,on grid,align=left] (b) {\texttt{B b = new B(a);}\\\texttt{b.getA();}};
  \node[above right=of a,on grid, align=left] (c) {\texttt{C c = new C(a);}\\\texttt{c.getA();}};

  % Arrows
  \draw[-{Latex[length=2mm]}] (b) -> (a);
  \draw[-{Latex[length=2mm]}] (c) -> (a);

  % Annotations
%   \node[below=0.5cm of a] (note) {Both `\texttt{b.getA()}' and `\texttt{c.getA()}' refer to the same `\texttt{a}' instance in memory};
% \node[below of=note] (note2) {Assume `\texttt{a}' is instantiated as `\texttt{new A()}' and points to address \texttt{0xffff2345}};

  % Equality Check Illustration (optional)
  % ... (Can be an inset showing assertEquals usage)
\end{tikzpicture}
\label{fig:memoryref}
\end{figure}

Comparing \ttt{b.getA().getS()} against \ttt{c.getA().getS()} is a comparison of two references to the same object, similar to the previous problem, right? Wrong! Recall that the \ttt{String} class overrides the \ttt{.equals} method implementation to compare strings for their content rather than their reference. Should we choose to compare the two strings for referential equality, we must use the \ttt{==} operator. In this case, we use \ttt{assertEquals} since the two strings are equal in content, but using \ttt{==} would also work because the strings are also equal in reference.

In the third line we change the value of the string inside \ttt{a} to be \ttt{"Hi!"}, which updates across all instances that point to \ttt{a}. Therefore, rerunning the same comparison as before still results in a true equality.

In the fifth line, we retrieve the \ttt{A} object instance pointed to by \ttt{B} and change its underlying string to be \ttt{"Howdy!"}. Rerunning the same test as before yet again results in a true equality. Because \ttt{b} points to the same \ttt{a} that \ttt{c} references, this change propagates across all references to \ttt{a}, even if we do not directly modify \ttt{a}.

We then declare a new instance of \ttt{B} named \ttt{b2}, which references the same \ttt{a} as before. If we check the value of \ttt{a} against the value of \ttt{a} inside \ttt{b2}, we of course get a true equality.

We immediately follow this comparison with one in which we compare \ttt{b} to \ttt{b2}. Because these are completely distinct object instantiations, the equality does not hold true.

Up next we reassign \ttt{b} to point to \ttt{b2}. This is a reassignment of a reference, not a reassignment of an object. Therefore if we check \ttt{b} against \ttt{b2} for equality, it is now trivially true.

Let's now test the \ttt{D} and \ttt{E} classes, which use arrays as instance variables.

\begin{cl}[]{Testing Object Mutation and Aliasing}
\begin{lstlisting}[language=MyJava]
import static Assertions.assertAll;
import static Assertions.assertEquals;

class ClassTester {

  @Test
  void testTwo() {
    D d = new D();
    E e = new E(42);
    E[] arrOfE = new E[10];
    for (int i = 0; i < arrOfE.length; i++) { arrOfE[i] = new E(i); }
    assert______(arrOfE[2], arrOfE[5]);
    assert______(arrOfE[2].getNumber(), arrOfE[5].getNumber());

    for (int i = 0; i < arrOfE.length; i++) { arrOfE[i] = e; }
    assert______(arrOfE[0], arrOfE[2]);
    assert______(arrOfE[0].getNumber(), arrOfE[2].getNumber());
    arrOfE[7].setNumber(102);
    assert______(arrOfE[0].getNumber(), arrOfE[2].getNumber());
  }
}
\end{lstlisting}
\end{cl}

The objects \ttt{d} and \ttt{e} are instantiated to types \ttt{D} and \ttt{E} respectively, the latter of which receives the integer \ttt{42} as an argument to the constructor. We follow this up with an array of ten \ttt{E} objects. The loop after instantiates each element of the array to a new \ttt{E} object with the integer \ttt{i} as an argument to the constructor.

So, what happens if we compare any arbitrary element $e$ against any other arbitrary element $e'$ such that $e \neq e'$? Because they are all instantiated to distinct instances of \ttt{E}, any equality comparison is false. We can extend this to retrieving the number inside each \ttt{E} object and comparing them. Because they are all distinct objects, and each \ttt{E} instance receives a different number, the equality is false.

The second for loop assigns each element of the array to the \ttt{e} object. Then, we can compare any arbitrary element against any other arbitrary element, and they will always be equal, since every element is a reference to the same memory reference. 

Thus, if we set the number of one arbitrary element, this change propagates to every other element in the list, because again, all references point to the same object.

\example{Some may question why we emphasize aliasing and mutation. When working with the Collections API and designing data structures, proper care must be taken to avoid undesired behavior. Consider what happens if we design a class \ttt{F} that stores a reference to a \ttt{List<Integer>} as an instance variable. Then, suppose we instantiate two distinct instances of \ttt{F}, namely $f_1$ and $f_2$, each of which receive the same list of numbers. If we then mutate the list somewhere inside of $f_1$, then the list stored as a reference inside $f_2$ is also changed.}

\begin{cl}[]{Testing List Mutation and Aliasing}
\begin{lstlisting}[language=MyJava]
import static Assertions.assertAll;
import static Assertions.assertEquals;

class ClassTester {

  @Test
  void testListAliasing() {
    List<Integer> ls = new ArrayList<>(List.of(1, 2, 3, 4, 5));
    F f1 = new F(ls);
    F f2 = new F(ls);
    f1.getList().set(2, 100);
    assertEquals(100, f1.getList().get(2));
    assertEquals(100, f2.getList().get(2));
  }
}
\end{lstlisting}
\end{cl}

\begin{cl}[]{Class Storing a List of Integers}
\begin{lstlisting}[language=MyJava]
import java.util.List;

class F {
  
  private final List<Integer> LS;
  
  public F(List<Integer> ls) { this.LS = ls; }
  
  public List<Integer> getList() { return this.LS; }
}
\end{lstlisting}
\end{cl}

\example{Recall the \ttt{LinkedList} class from Chapter~\ref{chapter-arrays-collections}. If you have ever wondered how it works under the hood, now is the time to find out! We will design a \emph{doubly-linked list} data structure that stores arbitrarily-typed elements.}

First, remember the structure of a linked list: they are composed of nodes, which hold the data and a pointer to the next element in the chain/sequence. These types of linked lists are \emph{singly-linked}, because nodes only refer to the successive element. In contrast, our class models a doubly-linked list, since its nodes point to their predecessor and their successor. 

We need a generic class that stores references to the first and last elements of the list. Let's create the \ttt{DoublyLinkedList} class to receive a type parameter \ttt{T} and store the first and last nodes as instance variables. It's important to realize that, whoever uses this class will not be exposed to the innards of the class, i.e., how the links are established/constructed/altered/removed. We wish to preserve the idea of encapsulation, after all.

We run into an eminent problem when declaring the types of the instance variables: what should they be? We need to design a class that encapsulates the value of the node, and holds references to its previous and successor nodes. Some programmers may consider designing a separate \ttt{.java} file for this class, but remember the encapsulation methodology: nobody outside of this class should even be aware that nodes exist in the first place. So, we can create a private and static \ttt{Node<T>} class, which is local to the definition of \ttt{DoublyLinkedList}. A privatized class can only ever be static, because it does not make sense to say that a private class definition belongs to an arbitrary instance of the class in which it resides. We also override the \ttt{toString} to output the underlying stringified data of the node.

\begin{cl}[]{Doubly-Linked List Class}
\begin{lstlisting}[language=MyJava]
class DoublyLinkedList<T> {

  private static class Node<T> {

    private T value;
    private Node<T> prev;
    private Node<T> next;

    private Node(T value) { 
      this.value = value; 
    }

    @Override
    public String toString() {
      return this.value.toString();
    }
  }

  private Node<T> head;
  private Node<T> tail;

  public DoublyLinkedList() {
    this.last = this.first = null;
  }
}
\end{lstlisting}
\end{cl}

Notice that, in the constructor of \ttt{DoublyLinkedList}, we assign the first and last references to each other, which both point to \ttt{null}. This is because, when the list is empty, there is no first or last element.

To test the methods that we are about to design, we will override the \ttt{toString} method (of \ttt{DoublyLinkedList}) to print the elements inside brackets, separated by commas and a space. To traverse over the list, however, we should use a custom-defined \ttt{Iterator}, which will be its own localized class definition. We have seen iterators before, but until now we have not implemented one on our own. The idea is, fortunately, very simple: we keep track of the current node, and upon calling \ttt{hasNext}, we return whether or not the node is \ttt{null}. Similarly, invoking \ttt{next} returns the value of the stored node and moves the pointer forward via the ``next'' instance. Finally, we create the \ttt{.iterator} method, which returns an instance of the iterator superclass. We do not want to expose how this particular iterator works, because the caller does not need to be aware of this logic; they are only concerned with iterating over the structure, in this case, a doubly-linked list.

\begin{cl}[]{Doubly-Linked List Iterator}
\begin{lstlisting}[language=MyJava]
import java.util.Iterator;

class DoublyLinkedList<T> {
  // ... previous code omitted ...

  public Iterator<T> iterator() {
    return new DoublyLinkedListIterator<>(this.first);
  }

  private static class DoublyLinkedListIterator<T> implements Iterator<T> {
    
    private Node<T> current;

    private DoublyLinkedListIterator(Node<T> first) {
      this.current = first;
    }

    @Override
    public boolean hasNext() { 
      return this.current != null; 
    }

    @Override
    public T next() {
      T value = this.current.value;
      this.current = this.current.next;
      return value;
    }
  }
}
\end{lstlisting}
\end{cl}

Using the iterator in \ttt{toString} is straightforward: we have a while loop that continues until no more elements are present. We complete two tasks at the same time by having an iterator, which then makes subsequent traversals over the list easier.

Now we can write methods to add, retrieve, and remove elements from the list. To add an elemenet, we need to take the links of \ttt{first} and \ttt{last}, and reassign them accordingly to remain consistent with our doubly-linked list property. If the list is empty, then we just have to assign the new node $n$ to both the \ttt{first} and \ttt{last} references. Otherwise, we set the ``next'' pointer of \ttt{last} to $n$, and set the ``previous'' pointer of $n$ to \ttt{last}. 

\begin{cl}[]{Testing the Doubly-Linked List Add Method}
\begin{lstlisting}[language=MyJava]
import static Assertions.assertAll;
import static Assertions.assertEquals;

class DoublyLinkedListTester {

  @Test
  void testAdd() {
    DoublyLinkedList<Integer> list = new DoublyLinkedList<>();
    assertAll(
      () -> assertEquals("[]", list.toString()),
      () -> list.add(1),
      () -> list.add(2),
      () -> list.add(3),
      () -> assertEquals("[1, 2, 3]", list.toString()),
      () -> list.add(4),
      () -> list.add(1),
      () -> list.add(5),
      () -> assertEquals("[1, 2, 3, 4, 1, 5]", list.toString()));
  }
}
\end{lstlisting}
\end{cl}

\begin{cl}[]{Doubly-Linked List Add Method}
\begin{lstlisting}[language=MyJava]
class DoublyLinkedList<T> {
  // ... previous code omitted ...

  /**
  * Adds a new node to the end of the list.
  * @param data The data to be stored in the new node.
  */
  public void add(T data) {
    Node<T> newNode = new Node<>(data);

    // If the list is empty, make the new node the first and last node.
    if (this.first == null) {
      this.first = newNode;
    } else {
      // Otherwise, add the new node to the end of the list.
      newNode.prev = this.last;
      this.last.next = newNode;
    }
    this.last = newNode;
  }
}
\end{lstlisting}
\end{cl}

Retrieving an element is trivial, as it's just a matter of traversing over the list and returning the data at the index of a node. If the index is out of bounds, we return an empty \ttt{Optional}.\footnote{It is, in general, a better idea to use exceptions, but we have not covered them yet.}

\begin{cl}[]{Testing the Doubly-Linked List Get Method}
\begin{lstlisting}[language=MyJava]
import static Assertions.assertAll;
import static Assertions.assertEquals;

class DoublyLinkedListTester {

  @Test
  void testGet() {
    DoublyLinkedList<Integer> list = new DoublyLinkedList<>();
    assertAll(
      () -> assertEquals(Optional.empty(), list.get(0)),
      () -> list.add(50),
      () -> list.add(25),
      () -> list.add(100),
      () -> assertEquals(Optional.of(50), list.get(0)),
      () -> assertEquals(Optional.of(25), list.get(1)),
      () -> assertEquals(Optional.of(100), list.get(2)),
      () -> assertEquals(Optional.empty(), list.get(3)),
      () -> list.add(1000),
      () -> list.add(10000),
      () -> list.add(50),
      () -> assertEquals(Optional.of(1000), list.get(3)),
      () -> assertEquals(Optional.of(10000), list.get(4)),
      () -> assertEquals(Optional.of(50), list.get(5)),
      () -> assertEquals(Optional.empty(), list.get(6)));
  }
}
\end{lstlisting}
\end{cl}

\begin{cl}[]{Doubly-Linked List Get Method}
\begin{lstlisting}[language=MyJava]
import java.util.Optional;

class DoublyLinkedList<T> {
  // ... previous code omitted ...

  /**
   * Returns the element at a given index as an Optional.
   * @param idx index to retrieve.
   * @return Optional.empty() if the index is out of bounds,
   *         the data at that node's index otherwise.
   */
  public Optional<T> get(int idx) {
    Node<T> curr = this.first;
    int i = 0;
    while (curr != null && i < idx) {
      curr = curr.next;
      i++;
    }
    return idx >= 0 && curr != null
            ? Optional.of(curr.data)
            : Optional.empty();
  }
}
\end{lstlisting}
\end{cl}

Finally we arrive at element removal, which is not as cut-and-dry. We want to pass the element-to-remove (compared via \ttt{equals}), but we need to adjust the pointers accordingly. In particular, there are four cases to consider:

\begin{enumerate}[label=(\alph*)]
  \item If the element-to-remove $e$ is the first of the list, then its successor is now the first. Its previous pointer is adjusted to now point to \ttt{null}.
  \item If the element-to-remove $e$ is the last of the list, then its predecessor is now the last. Its next pointer is adjusted to now point to \ttt{null}.
  \item If the element to remove $e$ is neither the first nor the last, we retrieve its previous node $p$, its next node $n$, and assign $p_\emph{next} = n$, and $n_\emph{prev} = p$. This, in effect, ``delinks'' $e$ from the list, which gets consumed by the garbage collector.
  \item If the element-to-remove $e$ is not in the list, do nothing.
\end{enumerate}

\begin{cl}[]{Testing the Doubly-Linked List Remove Method}
\begin{lstlisting}[language=MyJava]
import static Assertions.assertAll;
import static Assertions.assertEquals;

class DoublyLinkedListTester {

  @Test
  void testRemove() {
    DoublyLinkedList<Integer> list = new DoublyLinkedList<>();
    assertAll(
      () -> list.add(50),
      () -> list.add(25),
      () -> list.add(100),
      () -> list.remove(50),
      () -> assertEquals("[25, 100]", list.toString()),
      () -> list.remove(100),
      () -> assertEquals("[25]", list.toString()),
      () -> list.remove(25),
      () -> assertEquals("[]", list.toString()),
      () -> list.remove(25),
      () -> assertEquals("[]", list.toString()));
  }
}
\end{lstlisting}
\end{cl}

\begin{cl}[]{Doubly-Linked List Remove Method}
\begin{lstlisting}[language=MyJava]
class DoublyLinkedList<T> {
  // ... previous code omitted ...

  /**
   * Removes an element from the linked list, if it exists.
   * @param data value to be removed, compared via .equals.
   */
  public void remove(T data) {
    Node<T> curr = this.first;
    while (curr != null) {
      if (curr.data.equals(data)) { // Case 1: if it's the first.
        if (curr == this.first) {
          curr.next = this.first.next;
          this.first = curr.next;
        } else if (curr == this.last) { // Case 2: if it's the last.
          curr.prev.next = null;
          this.last = curr.prev;
        } else { // Case 3: if it's anything else.
          curr.prev.next = curr.next;
          curr.next.prev = curr.prev;
        }
        break;
      } else { curr = curr.next; }
    }
  }
}
\end{lstlisting}
\end{cl}

\example{Some programming languages do not come standard with data structures such as \ttt{Map} or, if they do, they are cumbersome to utilize. A substitute for the common mapping data structure is called an \emph{association list}, originating with the Lisp programming language. Its desired purpose is nearly identical to that of a map, but with worse performance implications. In this example we will design such a structure, as if \ttt{Map} did not exist in Java.}

Associations lists, as their name implies, associate values to other values, just like a map. In dynamically-typed languages, e.g., Scheme, association lists accept any type as their key and any type as their value. Therefore, we could have an association list that maps a string to an integer, or an integer to a list of strings, and so on. Should we want to use truly arbitrary types in the list, we can assign \ttt{Object} to both key and value types.

Our association list will support several methods that are related to their functional programming eqiuvalents. In particular, we want a \ttt{lookup} method to retrieve the associated value of some element and an an \ttt{extend} method to add a new association. Note that the \ttt{extend} method will, rather than modifying the current association list, return a new association list with the new association added. This is because we want to preserve the idea of immutability, which is a common theme in functoinal programming. Association lists, therefore, need to have a ``parent'' pointer to keep track of those associations in the list that we extend from.\footnote{In the next section on abstract classes and interpreters, we will revisit this idea in greater detail.} We will also override the \ttt{toString} method to print the associations in a readable format. 

\begin{cl}[]{Association List Tester}
\begin{lstlisting}[language=MyJava]
import static Assertions.assertAll;
import static Assertions.assertEquals;

class AssociationListTester {

  @Test
  void testAssociationList() {
    AssociationList<String, Integer> list = new AssociationList<>();
    assertAll(
      () -> assertEquals("[]", list.toString()),
      () -> list = list.extend("a", 1),
      () -> assertEquals("[(a, 1)]", list.toString()),
      () -> list = list.extend("b", 2),
      () -> assertEquals("[(b, 2), (a, 1)]", list.toString()),
      () -> list = list.extend("c", 3),
      () -> assertEquals("[(c, 3), (b, 2), (a, 1)]", list.toString()),
      () -> assertEquals(Optional.of(3), list.lookup("c")),
      () -> assertEquals(Optional.of(2), list.lookup("b")),
      () -> assertEquals(Optional.of(1), list.lookup("a")),
      () -> assertEquals(Optional.empty(), list.lookup("d")));
  }
}
\end{lstlisting}
\end{cl}

\begin{cl}[]{Association List Class}
\begin{lstlisting}[language=MyJava]
import java.lang.StringBuilder;
import java.util.Optional;

class AssociationList<K, V> {
  
  private final K key;
  private final V value;
  private final AssociationList<K, V> parent;
  
  public AssociationList() {
    this.key = null;
    this.value = null;
    this.parent = null;
  }

  private AssociationList(K key, V value, AssociationList<K, V> parent) {
    this.key = key;
    this.value = value;
    this.parent = parent;
  }
  
  /**
   * Returns the value associated with a given key.
   * @param key the key to lookup.
   * @return the value associated with the key, if it exists.
   */
  public Optional<V> lookup(K key) {
    AssociationList<K, V> curr = this;
    while (curr != null) {
      if (curr.key.equals(key)) {
        return Optional.of(curr.value);
      } else {
        curr = curr.parent;
      }
    }
    return Optional.empty();
  }
  
  /**
   * Adds a new association to the list.
   * @param key the key to associate.
   * @param value the value to associate.
   * @return a new association list with the new association.
   */
   public AssociationList<K, V> extend(K key, V value) {
    return new AssociationList<>(key, value, this);
   }
  
  @Override
  public String toString() {
    StringBuilder sb = new StringBuilder("[");
    AssociationList<K, V> curr = this;
    while (curr != null) {
      sb.append(String.format("(%s, %s)", curr.key, curr.value));
      curr = curr.parent;
      if (curr != null) { sb.append(", "); }
    }
    sb.append("]");
    return sb.toString();
  }
}
\end{lstlisting}
\end{cl}

Each association list in our representation stores exactly one association. Each time we extend the association, we create a new list that points to the previous list. This is a very inefficient way to store associations, but it is a common functional way of extending bindings in a programming language, most often in simple interpreted languages.