\setcounter{excounter}{1}
\setcounter{examplecounter}{1}
\chapter{Testing and Java Basics}
\label{chapter-testingandjava} % Always give a unique label
% use \chaptermark{}
% to alter or adjust the chapter heading in the running head

\abstract*{This chapter introduces readers to the basics of Java programming and testing. We will cover the fundamentals of Java syntax, including variables and data types. At the same time, we discuss the importance of test-driven development and how to write test cases in Java for methods. Our emphasis on testing makes work of the fact that testing is a critical part of writing correct and robust software. Test-driven development, in particular, is utilized as a ``design recipe,'' of sorts, to guide programmers to solve a task. }

\abstract{This chapter introduces readers to the basics of Java programming and testing. We will cover the fundamentals of Java syntax, including variables and data types. At the same time, we discuss the importance of test-driven development and how to write test cases in Java for methods. Our emphasis on testing makes work of the fact that testing is a critical part of writing correct and robust software. Test-driven development, in particular, is utilized as a ``design recipe,'' of sorts, to guide programmers to solve a task. }

\input{author/chapter-testingandjava/chapter-testingandjava-preliminaries}
\input{author/chapter-testingandjava/chapter-testingandjava-strings}
\input{author/chapter-testingandjava/chapter-testingandjava-stdiorand}
\newpage
\input{author/chapter-testingandjava/chapter-testingandjava-exercises}