\setcounter{excounter}{1}
\setcounter{examplecounter}{1}
\chapter{Conditionals, Recursion, and Loops}
\label{chapter-crl} % Always give a unique label
% use \chaptermark{}
% to alter or adjust the chapter heading in the running head

\abstract*{This chapter describes three core concepts in programming to readers: conditionals, recursion, and loops. We explain the Java syntax and intracacies surrounding \ttt{if}, \ttt{else if}, and \ttt{else}, as well as the illustrious \ttt{switch/case} statement. Repetition joins the discussion with recursion as a means of breaking down a problem into easier-to-solve smaller problems. We conclude with an introduction to loops via a translation pipeline from tail recursion, and consequently demonstrate that loops can be used as a substitute for any recursive algorithm, and may in fact be simpler for certain problems.}

\abstract{This chapter reintroduces three core concepts in programming to readers: conditionals, recursion, and loops. We explain the Java syntax and intracacies surrounding \ttt{if}, \ttt{else if}, and \ttt{else}, as well as the illustrious \ttt{switch/case} statement. Repetition joins the discussion with recursion as a means of breaking down a problem into easier-to-solve smaller problems. We conclude with an introduction to loops via a translation pipeline from tail recursion, and consequently demonstrate that loops can be used as a substitute for any recursive algorithm, and may in fact be simpler for certain problems.}

\input{author/chapter-crl/chapter-crl-conditionals}
\input{author/chapter-crl/chapter-crl-recursion}
\input{author/chapter-crl/chapter-crl-loops}
\newpage
\input{author/chapter-crl/chapter-crl-exercises}