\setcounter{excounter}{1}
\setcounter{examplecounter}{1}
\chapter{Advanced Object-Oriented Programming}
\label{chapter-advanced-oop} % Always give a unique label
% use \chaptermark{}
% to alter or adjust the chapter heading in the running head

\abstract*{Chapter~\ref{chapter-classes} demonstrated the potential of classes and the basics of object-oriented programming. In this chapter we expand on the details by describing interfaces and inheritance. Interfaces ascribe behaviors and characteristics to classes, whereas inheritance denotes ``IS-A'' relationships across classes. We conclude the chapter by demoing a practical application of object-oriented programming and class design by writing a small programming language.}

\abstract{Chapter~\ref{chapter-classes} demonstrated the potential of classes and the basics of object-oriented programming. In this chapter we expand on the details by describing interfaces and inheritance. Interfaces ascribe behaviors and characteristics to classes, whereas inheritance denotes ``IS-A'' relationships across classes. We conclude the chapter by demoing a practical application of object-oriented programming and class design by writing a small programming language.}

\input{author/chapter-advanced-oop/chapter-advanced-oop-interfaces}
\input{author/chapter-advanced-oop/chapter-advanced-oop-inheritance}
\input{author/chapter-advanced-oop/chapter-advanced-oop-abstract-classes}
\newpage
\input{author/chapter-advanced-oop/chapter-advanced-oop-exercises}